---
title: "Milwaukee Bucks Shooting Report, 2021-2026"
author: "Jacob Johnson"
date: "2026-03-28"
output: 
  html_document : default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
```

```{r message=FALSE, warning=FALSE, include=FALSE, results='hide', fig.show='hide'}
#installing need stat lib
#install.packages("hoopR")

#calling needed libs
library(hoopR)
library(hexbin)
library(ggplot2)
library(tidyverse)
```

```{r message=FALSE, warning=FALSE, include=FALSE, results='hide', fig.show='hide'}
##WORKING WITH THE DATA, MANIPULATING, EDA, ETC.
#loading data 'play by play'
nba_pbp = load_nba_pbp(2021:hoopR::most_recent_nba_season())


#pulling year out of the 'game_date' column
nba_pbp = nba_pbp %>% 
  mutate(year = year(ymd(game_date)))
unique(nba_pbp$year)    #checking

#filtering only mke buck's plays
bucks_pbp = nba_pbp %>% 
  filter((home_team_id == 15 | away_team_id == 15), #choosing team
         shooting_play == TRUE, #only shooting plays (no steals, etc.)
         )
bucks_pbp = bucks_pbp %>% 
  filter(team_id==15) #only bucks plays



##Mutating main data frame to fix plotting issues
#preping coordinates for plotting
shots_halfcourt = bucks_pbp %>%
  mutate(
    #swapping x and y axes
    x_raw = coordinate_y,
    y_raw = coordinate_x,
    
    #folding courting from full to half court
    y_half = ifelse(y_raw > 0, -y_raw, y_raw),
    
    #changing x range
    x_half = x_raw + 25,
    
    #changing y range
    y_half = y_half + 44.75,
    y_half = y_half * (47 / 44.75)
  )
#adding a col to determine if a shot was made or not (1=make, 0=not) (mainly for counts if needed)
shots_halfcourt = shots_halfcourt %>% 
  mutate(shot_made = as.numeric(score_value > 0))

#create col for legend use
shots_halfcourt$Result = factor(
  ifelse(shots_halfcourt$score_value > 0, "Make", "Miss"),
  levels = c("Make", "Miss")
)
#changing season variable to a factor so facet_wrap() graph subtitles work
shots_halfcourt$season = as.factor(shots_halfcourt$season)
```

```{r message=FALSE, warning=FALSE, include=FALSE, results='hide', fig.show='hide'}
##DRAWING BASKETBALL COURT

#Court creation Idea from 'https://www.rpubs.com/nMenai21/1363086', court is not exact I made some changes
#Change example was adding an arc function instead of estimating

###plotting shots

#Function for getting arc data (for plotting)
# Inputs: xc=x coord of center pt. yc=y coord of center pt. r=radius of arc,
#start_deg=starting degree of arc, end_deg=ending degree of arc, n=# for smoothness of the curve (higher #==more smooth)
# Outputs: data frame of points for arc for plotting
get_arc = function(xc, yc, r, start_deg, end_deg, n = 100) {
  theta = seq(start_deg * pi / 180, end_deg * pi / 180, length.out = n)
  data.frame(x = xc + r * cos(theta), y = yc + r * sin(theta))
}

# Curved line data frames
#3-Point Line (Calculated angle limit for the side-court clip)
angle_limit = asin((14 - 4.5) / 23.75) * 180 / pi
three_pt_arc = get_arc(25, 4.5, 23.75, angle_limit, 180 - angle_limit)

#Free Throw Circle (Top solid, Bottom dashed)
ft_top = get_arc(25, 19, 6, 0, 180)
ft_bottom = get_arc(25, 19, 6, 180, 360)

#Restricted Area (Curved part under/near the hoop)
restricted = get_arc(25, 4.5, 4, 0, 180)

#Half Court Circle
half_court = get_arc(25, 47, 6, 180, 360)

##COLORS
bucks_green = "#00471B" #buck green
bucks_cream = "#EEE1C6" #buck cream
court_wood  = "#E9D2A1" #Tan/Wood color for the floor



#Creating the half court
court = ggplot() +
  coord_fixed(xlim = c(0, 50), ylim = c(0, 47)) +
  
  #Coloring the background out of bounds bucks_green
  theme_void() +
  theme(plot.background = element_rect(fill=bucks_green, color=NA)) +
  
  #Out of bounds & Paint lines
  geom_rect(aes(xmin=0, xmax=50, ymin=0, ymax=47), fill=court_wood, color="black", linewidth=1) + #out of bounds box
  geom_rect(aes(xmin=17, xmax=33, ymin=0, ymax=19), fill=bucks_green, color=bucks_cream,
            linewidth=1) + #paint box

  #3-Point Line (line side segments & top arc)
  geom_segment(aes(x=3, y=0, xend=3, yend=14)) + #left line
  geom_segment(aes(x=47, y=0, xend=47, yend=14)) + #right line
  geom_path(data = three_pt_arc, aes(x, y)) + #top curve

  #Free Throw Lines (curves)
  geom_path(data = ft_top, aes(x, y)) + #ft top curve
  geom_path(data = ft_bottom, aes(x, y), linetype = "dashed") + #ft bottom dashed curve

  #Half court curve
  geom_path(data = half_court, aes(x, y)) +
  
  #Restricted area curve
  geom_path(data = restricted, aes(x, y)) +

  #Rim & Backboard
  geom_segment(aes(x=22, y=4, xend=28, yend=4), linewidth=1.5) +
  geom_point(aes(x=25, y=4.5), color="black", size=4) +
  
  labs(title = "Milwaukee Bucks ~ Cream City Court") +
  theme(plot.title = element_text(color="white", hjust=0.5, size=18, face="bold"))



#redoing court to make it a list of layers, instead of 1 (so it dosen't messed facet_wrap())
court_layers <- list(
  coord_fixed(xlim = c(0, 50), ylim = c(0, 47)),
  theme_void(),
  theme(
    plot.background = element_rect(fill = bucks_green, color = NA),
    plot.title = element_text(color = "white", hjust = 0.5, size = 18, face = "bold"),
    legend.text = element_text(color = 'white'),
    legend.title = element_text(color = 'white'),
    strip.text = element_text(color = "white", face = "bold", size = 12) # For facet labels
  ),
  
  #Base Floor
  geom_rect(aes(xmin=0, xmax=50, ymin=0, ymax=47), fill=court_wood, color="black", linewidth=1),
  
  #Paint - Note: using inherit.aes = FALSE so it doesn't look for 'Result' or 'Season'
  geom_rect(aes(xmin=17, xmax=33, ymin=0, ymax=19), fill=bucks_green, color=bucks_cream, linewidth=1, inherit.aes = FALSE),

  #3-Point Lines
  geom_segment(aes(x=3, y=0, xend=3, yend=14), inherit.aes = FALSE),
  geom_segment(aes(x=47, y=0, xend=47, yend=14), inherit.aes = FALSE),
  geom_path(data = three_pt_arc, aes(x, y), inherit.aes = FALSE),

  #FT and Restricted
  geom_path(data = ft_top, aes(x, y), inherit.aes = FALSE),
  geom_path(data = ft_bottom, aes(x, y), linetype = "dashed", inherit.aes = FALSE),
  geom_path(data = half_court, aes(x, y), inherit.aes = FALSE),
  geom_path(data = restricted, aes(x, y), inherit.aes = FALSE),

  #Rim & Backboard
  geom_segment(aes(x=22, y=4, xend=28, yend=4), linewidth=1.5, inherit.aes = FALSE),
  geom_point(aes(x=25, y=4.5), color="black", size=4, inherit.aes = FALSE)
)
```

```{r message=FALSE, warning=FALSE, include=FALSE, results='hide', fig.show='hide'}

###plotting shots on courts (over seasons/years with facet_wrap)

#Plotting makes and misses 
#Build wrapped jitter plot across seasons
makes_and_misses_bySeason = ggplot(shots_halfcourt, aes(x = x_half, y = y_half)) +
  court_layers + # Add the list of court elements
  geom_jitter(
    aes(shape = Result, color = Result),
    size = 1.8, 
    alpha = 0.6
  ) +
  scale_shape_manual(values = c("Make" = 19, "Miss" = 4), name = "Result") +
  scale_color_manual(values = c("Make" = "green", "Miss" = "red"), name = "Result") +
  guides(shape = guide_legend(override.aes = list(color = c("green", "red")))) +
  facet_wrap(~season) +
  labs(title = "Milwaukee Bucks ~ Makes and Misses by Season")

#Shot Density by Season
#Build wrapped jitter plot across seasons
hex_bin_shot_count_bySeason = ggplot(shots_halfcourt, aes(x = x_half, y = y_half)) +
  court_layers +  # Add court first
  geom_hex(
    bins = 30,
    color = "white",
    alpha = 0.8
  ) +
  scale_fill_viridis_c(
    option = "viridis", 
    limits = c(1, 50),
    name = "Shot Count"
  ) +
  facet_wrap(~season) +
  labs(title = "Milwaukee Bucks ~ Shot Density by Season")

#Shooting Efficiency by Season
#Build wrapped jitter plot across seasons
hex_bin_fg_percentage_bySeason = ggplot(shots_halfcourt, aes(x = x_half, y = y_half, z = shot_made)) +
  court_layers +  # Add court first
  stat_summary_hex(
    fun = mean,
    bins = 30,
    color = "white",
    alpha = 0.8
  ) +
  scale_fill_viridis_c(
    option = "viridis",
    labels = scales::percent,
    limits = c(0, 1),
    name = "FG%"
  ) +
  facet_wrap(~season) +
  labs(title = "Milwaukee Bucks ~ Shooting Efficiency by Season")
```


```{r, echo=FALSE}
#displaying 'makes and misses' and 'shot density'
makes_and_misses_bySeason
hex_bin_shot_count_bySeason
```

## Findings

- The restricted area in the paint over the years has stayed being the team's highest shot volume area. 
- Beyond the three point line although spread out is next area with high density of shots taken.
- Corner shots both in front and behind the three point line have also increased over time. 


## More In Depth

### Interior & Paint Frequency

- Looking at the shot density plot which uses heat based color scheme (warm colors mean more shots), we see from 2021-2026 the plot consistently shows a brighter/warmer spot in the paint near or in the restricted area of the paint. This shows how the team has stayed consistent in favoring the paint as their location of shot selection. Also looking at the makes and misses graph we see that majority of the shots taken near the restricted area have been consistently made over the years.

### Changes three points attempts

- Visually from both graphs we can see a change in the amount and locations of three point attempts taken. Looking at 2021 the three point attempts were very close to the three point line itself. By 2025 and 2026 best seen in the density graph the amount of hexes has grown deeper and deeper out from the three point line.

### Corner and Mid-range shots

- First touching on change of mid-range shots which are shots inside the three point line and outside the paint. Over time as seen in both graphs we see have seen a decline and growing absence of shots in front of the three point line, shown by either the gaps of points or density hexes. Digging deeper, while this is true for shots in front of the three point arc, the corners both inside the three point line and outside have grown in the number of attempts. Along with the number of attempts increasing looking at the makes and misses graph shows that shots taken in or near the corner have high amounts of makes. 


## Shot Efficiency around the court

```{r, echo=FALSE}
hex_bin_fg_percentage_bySeason
```


- This section will be diving deeper into **field goal percentage (FG%)**. By grouping shots into bins and calculating the the FG% for each seperate bin, we can see the team's overall shooting efficiency across the court. In the graph the color choices are relatively the same, for the shooting efficiency by season above the general idea is, the warmer the color the higher the FG%. Looking over the seasons we can see various changes in "hot" shot locations. Although this is true the area around the rim stays the highest concentration of high FG%.

- Another observation to look at is the change of mid-range shooting efficiency. The variance in mid-range is very inconsistent. Looking at the 2022 and 2023 seasons, there is an increased FG% on the right hand side of the court for mid-ranged shots. While looking at the 2024 season the mid-range FG% was still consistent in the right hand side, but there the high FG% also grew to the top of the key.

- The last key observation of the graph is the very consistent high FG% percentage in the right hand corner of the court.

